---
title: "PMDD Block Builder: Converting Excel into QSF Files"
output: html_document
params:
  excel_path: "scales.xlsx"      # path to your workbook
  seed: 1234                     # for reproducibility
  block_names: ["A","B","C"]     # change if you ever want 2 or 4 blocks
editor_options: 
  markdown: 
    wrap: 72
---

# A) How to use this template to convert your questionaire into a PMDD Qualtrics Survey

1.  Prepare your questionnaire Excel files as described in Section B.
2.  Run Section C in your R Markdown to load all required functions.
3.  Convert your Excel questionnaires (Section D) into PMDD-style
    blocks:
    -   choose how many items per factor should go into the X-block

    -   choose into how many forms (usually 3) the remaining items will
        be split

    -   the script automatically generates sub-blocks per scale (e.g. A
        BFI, X HEXACO, etc.)
4.  Import the generated TXT file into Qualtrics (Section E)
5.  Build the Survey Flow manually in Qualtrics following the
    instructions in Section F.
6.  Add demographic items to the X Block

# B) Required files in the working folder

### B1) scales.xlsx

This file contain the actual questionnaire items for all scales.

-   Filename must be exactly: scales.xlsx

-   One sheet per scale: The sheet name will become the scale identifier
    (e.g. BFI, PANAS, TIPI).

-   Required columns (case-insensitive)

    -   `factor` — the factor or subscale name/code\
    -   `item` — the item text to show in Qualtrics\
    -   `scale_level` — the name of the Likert scale you want to use
        (must match file B2)

-   **Optional columns** (kept and carried through automatically):

    -   `item_id` (unique within each sheet)\
    -   `reverse_keyed`\
    -   any other metadata columns you may want to include (for example:
        `language`, `block`, `order`, etc.)

**Example (sheet “BFI” inside scales.xlsx)**

| factor | item                                  | item_id | scale_level |
|:-------|:--------------------------------------|:--------|:------------|
| E      | Is talkative.                         | E1      | likert5     |
| E      | Is reserved.                          | E2R     | likert5     |
| A      | Tends to find fault with others.      | A1      | likert5     |
| A      | Is helpful and unselfish with others. | A2      | likert5     |
| …      | …                                     | …       | …           |

------------------------------------------------------------------------

### B2) scale_levels.xlsx

This file defines the response options for each Likert scale used in
scales.xlsx

-   Filename must be exactly scale_levels.xlsx

<!-- -->

-   **Required columns:**

    -   `scale_level` — the name used in scales.xlsx has to match the
        one used in scale_levels.xlsx (e.g. likert5)\
    -   `value` — the numeric value saved in Qualtrics\
    -   `label` — text shown to participants

**Example (file “scale_levels.xlsx”)**

| scale_level | value | label                      |
|:------------|:------|:---------------------------|
| likert5     | 1     | Strongly disagree          |
| likert5     | 2     | Disagree                   |
| likert5     | 3     | Neither agree nor disagree |
| likert5     | 4     | Agree                      |
| likert5     | 5     | Strongly agree             |
| likert7     | 1     | Never                      |
| likert7     | 2     | Rarely                     |
| likert7     | 3     | Sometimes                  |
| ...         | ...   | ...                        |

------------------------------------------------------------------------

# C) Preparing the conversion-functions

##C1) Packages

```{r setup, message=FALSE, warning=FALSE}
# install.packages("readxl")
library(readxl)

# install.packages("dplyr")
library(dplyr)

# install.packages("tidyr")
library(tidyr)

# install.packages("purrr")
library(purrr)

# install.packages("stringr")
library(stringr)

# install.packages("readr")
library(readr)

# install.packages("rlang")
library(rlang)

# install.packages("janitor")
library(janitor)


set.seed(params$seed)

excel_path  <- params$excel_path
block_names <- params$block_names
stopifnot(length(block_names) >= 2)
```

##C2) Function Identifying Scales

```{r}
## function identifying scales from excel sheet and creating data sets per sheet

datascales <- function(path){
  sheets <- readxl::excel_sheets(path)
  
  for (sheet in sheets) {
    df <- readxl::read_xlsx(path, sheet = sheet)
  
  assign(sheet, df, envir = .GlobalEnv)
  
  message("Created dataset: ", sheet)
  }
  scale_names <- tibble::tibble(scale_name = sheets)
  assign("scale_names", scale_names, envir = .GlobalEnv)
  
  message("Created 'scale_names' data frame.")
  
}

```

##C3) Function creating X Block

```{r}
## function creating X-block

create_xblock <- function(n) {
  # Validate n
  if (missing(n) || !is.numeric(n) || length(n) != 1 || n <= 0) {
    abort("`n` must be a single positive number (e.g., n = 2).")
  }
  
  # Check for scale_names data frame
  if (!exists("scale_names", envir = .GlobalEnv)) {
    abort("`scale_names` not found in the environment. Run `datascales()` first.")
  }
  
  scales_tbl <- get("scale_names", envir = .GlobalEnv)
  
  if (!"scale_name" %in% names(scales_tbl)) {
    abort("`scale_names` must have a column `scale_name` with sheet names.")
  }
  
  # Iterate over scales
  xblock <- purrr::map_dfr(scales_tbl$scale_name, function(sc) {
    if (!exists(sc, envir = .GlobalEnv)) {
      abort(paste0("Dataset '", sc, "' not found in the environment."))
    }
    
    df <- get(sc, envir = .GlobalEnv)
    
    # Basic column presence check
    required <- c("factor", "item")
    missing <- setdiff(required, names(df))
    if (length(missing) > 0) {
      abort(paste0("Dataset '", sc, "' is missing required columns: ",
                   paste(missing, collapse = ", ")))
    }
    
    # Select first n items per factor
    df %>%
      mutate(.row_order__ = dplyr::row_number()) %>%
      group_by(factor) %>%
      arrange(.row_order__, .by_group = TRUE) %>%
      slice_head(n = n) %>%
      ungroup() %>%
      select(-.row_order__) %>%
      mutate(scale = sc, .before = 1)
  })
  
  # Assign result to global environment
  assign("xblock", xblock, envir = .GlobalEnv)
  message("✅ Created `xblock` with ", nrow(xblock), " rows from ",
          length(unique(xblock$scale)), " scales.")
  
  invisible(xblock)
}

```

##C4) Function creating ABC Blocks

```{r}
## function creating ABC blocks

# --- Helper 1: Excel-like sequence: a, b, ..., z, aa, ab, ...
.make_block_keys <- function(n) {
  stopifnot(n >= 1)
  keys <- character(0)
  i <- 0
  while (length(keys) < n) {
    i <- i + 1
    # convert i to base-26 letters (1-indexed)
    x <- i
    s <- character(0)
    while (x > 0) {
      rem <- (x - 1) %% 26
      s <- c(letters[rem + 1], s)
      x <- (x - rem - 1) %/% 26
    }
    keys <- c(keys, paste0(s, collapse = ""))
  }
  keys[seq_len(n)]
}

# --- Helper 2: assign items of a tibble to blocks in round-robin (1..n, repeating)
# replace the helper with this version
.assign_round_robin <- function(df, n) {
  if (nrow(df) == 0) return(df %>% dplyr::mutate(.block_index = integer()))
  df %>%
    dplyr::mutate(.block_index = ((dplyr::row_number() - 1) %% n) + 1)
}


# --- Main: create_blocks -------------------------------------------------------
# x = number of items already taken for xblock (per factor)
# n = number of blocks to create (a, b, c, ...)
create_blocks <- function(x, n) {
  # validate
  if (missing(x) || !is.numeric(x) || length(x) != 1 || x < 0) {
    abort("`x` must be a single non-negative number.")
  }
  if (missing(n) || !is.numeric(n) || length(n) != 1 || n < 1) {
    abort("`n` must be a single positive number of blocks.")
  }
  x <- as.integer(x); n <- as.integer(n)

  # need scale_names
  if (!exists("scale_names", envir = .GlobalEnv)) {
    abort("`scale_names` not found. Run `datascales()` first.")
  }
  scales_tbl <- get("scale_names", envir = .GlobalEnv)

  # block names
  block_keys  <- .make_block_keys(n)
  block_names <- paste0(block_keys, "block")

  # init empty blocks
  blocks <- setNames(rep(list(tibble()), n), block_names)

  # per scale
  for (sc in scales_tbl$scale_name) {
    if (!exists(sc, envir = .GlobalEnv)) {
      abort(paste0("Dataset '", sc, "' not found in the environment."))
    }
    df <- get(sc, envir = .GlobalEnv)

    # required cols
    if (!all(c("factor","item") %in% names(df))) {
      abort(paste0("Dataset '", sc, "' must contain columns: factor, item."))
    }

    # keep original order
    df <- df %>% mutate(.row_order__ = row_number())

    # drop first x per factor; keep remaining
    df_alloc <- df %>%
      group_by(factor) %>%
      arrange(.row_order__, .by_group = TRUE) %>%
      mutate(.row_in_factor = row_number()) %>%
      filter(.row_in_factor > x) %>%
      select(-.row_in_factor) %>%
      ungroup()

    if (nrow(df_alloc) == 0) next

    # round-robin assign 1..n
    df_alloc <- .assign_round_robin(df_alloc, n)

    # append to each block by POSITION (1..n), not by split names
    for (i in seq_len(n)) {
      part <- df_alloc %>%
        filter(.block_index == i) %>%
        select(-.block_index) %>%
        arrange(.row_order__) %>%
        select(-.row_order__) %>%
        mutate(scale = sc, .before = 1)

      if (nrow(part) > 0) {
        blocks[[ block_names[i] ]] <- bind_rows(blocks[[ block_names[i] ]], part)
      }
    }
  }
  # assign to environment and return
  walk2(names(blocks), blocks, ~ assign(.x, .y, envir = .GlobalEnv))
  message("Created blocks: ", paste(names(blocks), collapse = ", "))
  invisible(blocks)
}

```

##C5) Function loading and validating scale_levels Excel

```{r}
# function reading and validating scale levels

read_scale_levels_file <- function(path) {
  sl <- readxl::read_xlsx(path) %>%
    mutate(
      scale_level = as.character(scale_level),
      value       = suppressWarnings(as.integer(value)),
      label       = as.character(label)
    ) %>%
    arrange(scale_level, value)
  
  # Required columns check
  required <- c("scale_level", "value", "label")
  missing  <- setdiff(required, names(sl))
  if (length(missing) > 0) {
    abort(paste0("Scale levels file is missing columns: ",
                 paste(missing, collapse = ", ")))
  }
  
  # Sanity checks
  if (any(is.na(sl$scale_level) | sl$scale_level == "")) {
    abort("Scale levels file has empty 'scale_level' entries.")
  }
  if (any(is.na(sl$value))) {
    abort("Scale levels file has missing or non-integer 'value' entries.")
  }
  if (any(is.na(sl$label) | sl$label == "")) {
    abort("Scale levels file has empty 'label' entries.")
  }
  
  # Split into named list: likert5, likert7, ...
  scale_levels <- split(sl, sl$scale_level) %>%
    map(~ select(.x, value, label) %>% arrange(value))
  
  # Store globally
  assign("scale_levels", scale_levels, envir = .GlobalEnv)
  message("✅ Created object 'scale_levels' with ",
          length(scale_levels), " scale definitions.")
  
  invisible(scale_levels)
}

```

##C6) Function creating TXT-Survey

```{r}
# Helper: setting up Questions as TXT
.default_choices <- tibble::tibble(
  value = 1:5,
  label = c(
    "Strongly disagree",
    "Disagree",
    "Neither agree nor disagree",
    "Agree",
    "Strongly agree"
  )
)

get_choices_for_row <- function(row, scale_levels = NULL) {
  if (!is.null(scale_levels) && "scale_level" %in% names(row)) {
    lvl <- as.character(row$scale_level[[1]])
    if (!is.na(lvl) && lvl %in% names(scale_levels)) {
      return(scale_levels[[lvl]])
    }
  }
  .default_choices
}

emit_question_txt <- function(row, scale_levels = NULL) {
  stopifnot(nrow(row) == 1)
  item_text <- as.character(row$item[[1]])
  choices_tbl <- get_choices_for_row(row, scale_levels)

  # Advanced-Format-Question
  c(
    "[[Question:MC]]",
    item_text,
    "[[Choices]]",
    choices_tbl$label,
    ""  # Leerzeile zum Trennen
  )
}

# Helper: Setting up Blocks as TXT
emit_block_txt <- function(block_df, block_title, scale_levels = NULL) {
  if (nrow(block_df) == 0) return(character(0))

  # Wir gehen in der gegebenen Reihenfolge durch die Items
  rows <- split(block_df, seq_len(nrow(block_df)))

  lines <- c(
    sprintf("[[Block:%s]]", block_title),
    ""
  )

  for (rw in rows) {
    lines <- c(lines, emit_question_txt(rw, scale_levels))
  }

  lines
}


# Helper: Setting up Sub-Blocks
emit_letter_subblocks_txt <- function(letter_df, letter, scale_levels = NULL) {
  if (is.null(letter_df) || nrow(letter_df) == 0) return(character(0))
  if (!"scale" %in% names(letter_df)) {
    stop("letter_df muss eine Spalte 'scale' haben (kommt aus datascales()/create_blocks()).")
  }

  # Nach Skala splitten
  split_list <- split(letter_df, letter_df$scale)

  lines <- character(0)

  for (sc_name in names(split_list)) {
    df_sc <- split_list[[sc_name]]

    # Blocktitel wählen – hier: "A BFI", "B HEXACO", "X BFI", ...
    block_title <- sprintf("%s %s", toupper(letter), sc_name)

    lines <- c(
      lines,
      emit_block_txt(df_sc, block_title, scale_levels),
      ""  # optional Leerzeile zwischen Subblöcken
    )
  }

  lines
}


# Setting up complete TXT Survey
build_txt_survey <- function(
  xblock,
  ablock = NULL,
  bblock = NULL,
  cblock = NULL,
  scale_levels = NULL,
  file = "PMDD_import.txt"
) {
  lines <- c(
    "[[AdvancedFormat]]",
    ""
  )

  # X-Subblöcke: X BFI, X CFI, ...
  if (!missing(xblock) && !is.null(xblock) && nrow(xblock) > 0) {
    lines <- c(lines, emit_letter_subblocks_txt(xblock, "X", scale_levels))
  }

  # A-Subblöcke: A BFI, A CFI, ...
  if (!is.null(ablock) && nrow(ablock) > 0) {
    lines <- c(lines, emit_letter_subblocks_txt(ablock, "A", scale_levels))
  }

  # B-Subblöcke: B BFI, B CFI, ...
  if (!is.null(bblock) && nrow(bblock) > 0) {
    lines <- c(lines, emit_letter_subblocks_txt(bblock, "B", scale_levels))
  }

  # C-Subblöcke: C BFI, C CFI, ...
  if (!is.null(cblock) && nrow(cblock) > 0) {
    lines <- c(lines, emit_letter_subblocks_txt(cblock, "C", scale_levels))
  }

  writeLines(lines, file, useBytes = TRUE)
  message("✅ TXT für Qualtrics geschrieben: ", normalizePath(file))
  invisible(file)
}


```

# D) Convert your Excel-Files into PMDD TXT-Files

```{r}
# ------------------------------------------------------------
# 1) Scales
# ------------------------------------------------------------

# Load your questionnaire sheets from "scales.xlsx".
datascales("scales.xlsx")

# create_xblock(n = ...)
# → n = number of items per *factor* that should go into the X-Block.
#   Usually: n = 1  (recommended: one anchor item per factor)
create_xblock(n = 1)

# create_blocks(x = ..., n = ...)
# → x = same number as above (items already taken into the X-Block)
# → n = number of forms you want to compare (e.g., 3 = A/B/C groups)
#
#   Typical PMDD: x = 1, n = 3
create_blocks(x = 1, n = 3)


# ------------------------------------------------------------
# 2) Scale Levels
# ------------------------------------------------------------

# Load your Likert scale definitions.
# Must match the "scale_level" column inside scales.xlsx.
read_scale_levels_file("scale_levels.xlsx")


# ------------------------------------------------------------
# 3) Generate TXT Survey
# ------------------------------------------------------------

# This creates the final TXT file that you import into Qualtrics.
#
# Arguments:
#   xblock = the X-Block created above
#   ablock, bblock, cblock = the randomized A/B/C blocks
#   file = name of the TXT survey file to be created
#
build_txt_survey(
  xblock       = xblock,
  ablock       = ablock,
  bblock       = bblock,
  cblock       = cblock,
  scale_levels = scale_levels,
  file         = "PMDD_Jingle_Fallacy_Study.txt"
)


```

# E) Import Your TXT-File into Qualtrics

**Option 1) Create a new project from your TXT file**

1.  Log in to Qualtrics

2.  Go to the Projects dashboard.

3.  Click Create Project (upper right).

4.  Choose Survey -\> Get Started

5.  Scroll down to Import a survey

6.  Select Upload a TXT File

7.  Upload: PMDD_Jingle_Fallacy_Study.txt

Qualtrics will now automatically create all blocks

No Survey Flow is created yet — this is expected.

**Option 2) Import into an existing project**

Use this option if you want to import the items into a project you
already created.

1.  Open your existing project in Qualtrics.

2.  Go to the Survey Tab.

3.  In the top menu, click Tools.

4.  Choose Import Survey.

5.  Upload: PMDD_Jingle_Fallacy_Study.txt

6.  Confirm the import.

Note: If the project already contains questions or blocks, the imported
blocks will simply be added below the existing ones.

# F) Create the PMDD Survey Flow inside Qualtrics

After importing your TXT file, your Qualtrics project contains all
blocks but no Survey Flow yet.

Now we will build the PMDD Flow manually following these steps.

Below a visual overview of the final structure you will build:

### F1) PMDD Survey Flow Overview (Diagram)

``` text
Start
│
├── Group X
│     ├── Embedded Data: Form = X
│     └── Randomizer (show all, random order)
│           ├── X – BFI
│           ├── X – HEXACO
│           └── X – … (one per scale)
│
└── Randomizer (show 1 of 3 forms: A, B, C)
      │
      ├── Group A
      │     ├── Embedded Data: Form = A
      │     └── Randomizer (show all, random order)
      │           ├── A – BFI
      │           ├── A – HEXACO
      │           └── A – … (one per scale)
      │
      ├── Group B
      │     ├── Embedded Data: Form = B
      │     └── Randomizer (show all, random order)
      │           ├── B – BFI
      │           ├── B – HEXACO
      │           └── B – … 
      │
      └── Group C
            ├── Embedded Data: Form = C
            └── Randomizer (show all, random order)
                  ├── C – BFI
                  ├── C – HEXACO
                  └── C – …
```

### F2) How to Build This Flow in Qualtrics (Step-by-Step)\*\*

**Step 1 — Open Survey Flow**

1.  Open your project

2.  Go to Survey → Survey Flow

3.  Delete anything already there (if present)

------------------------------------------------------------------------

**Step 2 — Add Group X**

1.  Click Add Below → Add a Group

2.  Rename group: Group X

3.  Inside Group X add:

-   Add a New Element → Embedded Data

-   Create field: `Form`

-   Set value: `X`

-   Add a New Element → Randomizer

**Configure the randomizer:**

-   Randomization Type: Show all elements

-   Randomize order: ✔ enabled

**Add all X-subblocks:**

Inside the randomizer, click Add Element and add each of these:

-   `X – BFI`

-   `X – HEXACO`

-   `X – …` (one per scale)

------------------------------------------------------------------------

**Step 3 — Add the outer randomizer for Forms A/B/C**

1.  Below Group X, click Add Below → Randomizer

2.  Set:

-   Randomization Type: Show 1 element

-   Evenly Present Elements: enabled

This ensures equal assignment to A, B, or C across participants.

------------------------------------------------------------------------

**Step 4 — Create each Form Group (A, B, C)**

You will repeat the following steps three times

Group A, Group B, and Group C

For each group:

1.  Inside the big randomizer, click Add Element → Group

2.  Rename to:\
    Group A (or B or C)

3.  Inside that group:\

    a)  Add Embedded Data

    -   Field: Form
    -   Value:`A` (or B or C)

    b)  Add a Randomizer

    -   Randomization Type: Show all elements
    -   Randomize Order: enabled

    c)  Add all subblocks for that form (e.g. for A):

    -   `A – BFI`
    -   `A – HEXACO`
    -   `A – …`\
        (one block per scale)

Repeat forGroup B and Group C.

------------------------------------------------------------------------

### F3) Checklist: Your Final Flow Should Look Like This

✔ One Group X at the top ✔ Inside it:

-   Embedded `Form = X`

-   Randomizer (show all, random order)

-   Subblocks: X–BFI, X–HEXACO, X–…

✔ Then one Randomizer (show 1 of 3, even presentation) ✔ Three groups:
A, B, C ✔ Each group contains:

-   Embedded `Form = A/B/C`

-   Randomizer (show all, random order)

-   All subblocks for that form

------------------------------------------------------------------------
